Generalized Plackett-Luce Likelihoods

### 2023-10-29 inst/jss4832.tex
various changes for JSS editorial policy

<<setup, cache=FALSE,echo=FALSE>>=
source("loadlibrary.R",echo=FALSE)  # to avoid stupid messages 
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(prompt = "R> ")
@

@@

<<>>=
LL <- hyper3()
LL[c(a = 1)] <- 1
LL[c(a = 3, b = 2, c = 1)] <- -1
LL
@

@@

<<>>=
loglik(c(a = 0.01, b = 0.01, c = 0.98), LL)
loglik(c(a = 0.90, b = 0.05, c = 0.05), LL)
@

@@

<<>>=
(H <- ordervec2supp3(c("a", "c", "b", "a", "a", "b")))
@

@@

<<maxexamp,cache=TRUE>>=
(mH <- maxp(H))
@

@@

<<testequality,cache=TRUE>>=
equalp.test(H)
@

@@

<<maxpaba,cache=TRUE>>=
maxp(ordervec2supp3(c("a", "b", "a")))
@

@@

<<>>=
a <- 1/2 # null
(S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
@

@@

<<>>=
pchisq(-2 * S_delta, df = 1, lower.tail = FALSE)
@

@@

\begin{figure}[htbp]
  \begin{center}
<<figaba, fig=TRUE>>=
a <- seq(from = 0, by = 0.005, to = 1)
S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
plot(a, S(a), type = 'b',xlab=expression(p[a]),ylab="support")
abline(h = c(0, -2))
abline(v = c(0.02438102, 0.9524271), col = 'red')
abline(v = sqrt(2) - 1)
@
\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$.}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
  \begin{center}
<<figabbabb,echo=FALSE,fig=TRUE>>=
f_aab <- function(a){a^2 / (1 + a)}
f_aba <- function(a){a * (1 - a) / (1 + a)}
f_baa <- function(a){(1 - a) / (1 + a)}
p <- function(f, ...){
  a <- seq(from = 0, by = 0.005, to = 1)
  points(a, f(a) / max(f(a)), ...)
  }
plot(0:1, 0:1, xlab = expression(p[a]), ylab = "Likelihood", type = "n")
p(f_aab, type = "l", col = "black")
p(f_aba, type = "l", col = "red")
p(f_baa, type = "l", col = "blue")
text(0.8,0.8,"AAB")
text(0.8,0.4,"ABA",col="red")
text(0.8,0.05,"BAA",col="blue")
abline(h = exp(-2), lty = 2)
@
\caption{Likelihood \label{ABAetc_likelihood} functions for
observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.  Horizontal dotted
line represents two units of support}
\end{center}
\end{figure}

@@

<<cache=TRUE>>=
maxp(ordervec2supp3(c("a", "b"), nonfinishers=c("a", "b")))
@

@@

<<define_xy_wilcox>>=
x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
@

@@

<<hyper3osdef,cache=TRUE>>=
names(x) <- rep("x", length(x))
names(y) <- rep("y", length(y))
(os <- names(sort(c(x, y))))
@

@@

<<hyper3xytest,cache=TRUE>>=
Hxy <- ordervec2supp3(os)
equalp.test(Hxy)
@

@@

\begin{figure}[htbp]
  \begin{center}
<<plotwilcoxlike,echo=FALSE,fig=TRUE,cache=TRUE>>=
a <- seq(from = 0.02, to = 0.8, len = 40)
L <- sapply(a, function(p){loglik(p, Hxy)})
plot(a, L - max(L), type = 'b',xlab=expression(p[a]),ylab="likelihood")
abline(h = c(0, -2))
abline(v = c(0.24))
abline(v=c(0.5), lty=2)
@
\caption{A\label{wilcox_likelihood} support function for the Bradley-Terry
strength $p_a$ of permeability at term.  The evaluate of 0.24 is
shown; and the two-units-of support credible interval, which does not
exclude $H_0\colon p_a=0.5$ (dotted line), is also shown.}
\end{center}
\end{figure}

@@

<<javelintable>>=
javelin_table
@

@@

<<label=converttosupp3, cache=TRUE>>=
javelin_vector <- attemptstable2supp3(javelin_table,
       decreasing = TRUE, give.supp = FALSE)
options(width = 60)
javelin_vector
@

@@

<<dothething2,cache=TRUE>>=
javelin <- ordervec2supp3(v = names(javelin_vector)[!is.na(javelin_vector)])
@

@@

<<setdigits,echo=FALSE>>=
options(digits = 3)
@

\begin{figure}[htbp]
  \begin{center}
<<testthejav, cache=TRUE, fig=TRUE>>=
(mj <- maxp(javelin))
dotchart(mj, pch = 16,xlab="Estimated Bradley-Terry strength")
@
\caption{Maximum \label{javelinbradleyterry} likelihood estimate for javelin throwers' Bradley-Terry strengths.}
\end{center}
\end{figure}

@@

<<havealook,echo=FALSE,cache=TRUE>>=
f <- function(s){
  jj <- specificp.test(javelin, "Vesely", s, n = 2)
  p <- jj$null_estimate
 return(c(log(p[6] / p[7]), jj$null_support))
}
Ves <- seq(from = 0.0199, to = 0.33, len = 16)
M <- sapply(Ves, f)
M[2,] <- M[2,] - max(M[2,])
rownames(M) <- c("logcontrast", "support")
@

\begin{figure}[htbp]
\begin{center}
<<plottheloglikcont,echo=FALSE,fig=TRUE,cache=TRUE>>=
colnames(M) <-  as.character(Ves)
plot(t(M), type = "b")
abline(h = c(0, -2))
abline(v = 0, lty = 2)
abline(v = log(0.32062833 / 0.11402735)) # these from mp
@
\caption{Profile \label{profliklogcont} likelihood for log-contrast
$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$.  Null
of $p_{\mbox{\tiny Vad}}=p_{\mbox{\tiny Ves}}$ indicated with a dotted
line, and two-units-of-support limit indicated with horizontal lines;
thus the null is not rejected.}
\end{center}
\end{figure}

@@

<<showconstructortable>>=
constructor_2021_table[, 1:9]
@

@@

<<maxpconstructor2021, cache=TRUE>>=
const2020 <- ordertable2supp3(constructor_2020_table)
const2021 <- ordertable2supp3(constructor_2021_table)
options(digits = 4)
maxp(const2020, n = 1)
maxp(const2021, n = 1)
@

@@

<<definecombinedlikelihoodfunction,cache=TRUE>>=
H <- (
      psubs(constructor_2020, "Merc", "Merc2020") +
      psubs(constructor_2021, "Merc", "Merc2021")
     )
@

@@

<<usesamep,cache=TRUE>>=
options(digits = 4)
samep.test(H, c("Merc2020", "Merc2021"))
@

@@

\bibliography{jss_hyper3}
\end{document}

>>>>

@@

\begin{Schunk}
\begin{Sinput}
R> LL <- hyper3()
R> LL[c(a = 1)] <- 1
R> LL[c(a = 3, b = 2, c = 1)] <- -1
R> LL
\end{Sinput}
\begin{Soutput}
log( (a=1)^1 * (a=3, b=2, c=1)^-1)
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> loglik(c(a = 0.01, b = 0.01, c = 0.98), LL)
\end{Sinput}
\begin{Soutput}
[1] -4.634729
\end{Soutput}
\begin{Sinput}
R> loglik(c(a = 0.90, b = 0.05, c = 0.05), LL)
\end{Sinput}
\begin{Soutput}
[1] -1.15268
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> (H <- ordervec2supp3(c("a", "c", "b", "a", "a", "b")))
\end{Sinput}
\begin{Soutput}
log( (a=1)^3 * (a=1, b=1)^-1 * (a=2, b=1)^-1 * (a=2, b=2)^-1 * (a=2,
b=2, c=1)^-1 * (a=3, b=2, c=1)^-1 * (b=1)^1 * (c=1)^1)
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> (mH <- maxp(H))
\end{Sinput}
\begin{Soutput}
         a          b          c 
0.21324090 0.08724824 0.69951086 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> equalp.test(H)
\end{Sinput}
\begin{Soutput}
	Constrained support maximization

data:  H
null hypothesis: a = b = c
null estimate:
        a         b         c 
0.3333333 0.3333333 0.3333333 
(argmax, constrained optimization)
Support for null:  -6.579251 + K

alternative hypothesis:  sum p_i=1 
alternative estimate:
         a          b          c 
0.21324090 0.08724824 0.69951086 
(argmax, free optimization)
Support for alternative:  -5.73209 + K

degrees of freedom: 2
support difference = 0.8471613
p-value: 0.42863 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> maxp(ordervec2supp3(c("a", "b", "a")))
\end{Sinput}
\begin{Soutput}
        a         b 
0.4142108 0.5857892 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> a <- 1/2 # null
R> (S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
\end{Sinput}
\begin{Soutput}
[1] -0.0290123
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> pchisq(-2 * S_delta, df = 1, lower.tail = FALSE)
\end{Sinput}
\begin{Soutput}
[1] 0.8096458
\end{Soutput}
\end{Schunk}

@@

\begin{figure}[htbp]
  \begin{center}
\begin{Schunk}
\begin{Sinput}
R> a <- seq(from = 0, by = 0.005, to = 1)
R> S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
R> plot(a, S(a), type = 'b',xlab=expression(p[a]),ylab="support")
R> abline(h = c(0, -2))
R> abline(v = c(0.02438102, 0.9524271), col = 'red')
R> abline(v = sqrt(2) - 1)
\end{Sinput}
\end{Schunk}
\includegraphics{jss_hyper3-figaba}
\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$.}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
  \begin{center}
\includegraphics{jss_hyper3-figabbabb}
\caption{Likelihood \label{ABAetc_likelihood} functions for
observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.  Horizontal dotted
line represents two units of support}
\end{center}
\end{figure}

@@

\begin{Schunk}
\begin{Sinput}
R> maxp(ordervec2supp3(c("a", "b"), nonfinishers=c("a", "b")))
\end{Sinput}
\begin{Soutput}
        a         b 
0.5857892 0.4142108 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
R> y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> names(x) <- rep("x", length(x))
R> names(y) <- rep("y", length(y))
R> (os <- names(sort(c(x, y))))
\end{Sinput}
\begin{Soutput}
 [1] "x" "y" "x" "x" "y" "y" "x" "y" "y" "x" "x" "x" "x" "x" "x"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> Hxy <- ordervec2supp3(os)
R> equalp.test(Hxy)
\end{Sinput}
\begin{Soutput}
	Constrained support maximization

data:  Hxy
null hypothesis: x = y
null estimate:
  x   y 
0.5 0.5 
(argmax, constrained optimization)
Support for null:  -27.89927 + K

alternative hypothesis:  sum p_i=1 
alternative estimate:
        x         y 
0.2401539 0.7598461 
(argmax, free optimization)
Support for alternative:  -26.48443 + K

degrees of freedom: 1
support difference = 1.414837
p-value: 0.09253716 
\end{Soutput}
\end{Schunk}

@@

\begin{figure}[htbp]
  \begin{center}
\includegraphics{jss_hyper3-plotwilcoxlike}
\caption{A\label{wilcox_likelihood} support function for the Bradley-Terry
strength $p_a$ of permeability at term.  The evaluate of 0.24 is
shown; and the two-units-of support credible interval, which does not
exclude $H_0\colon p_a=0.5$ (dotted line), is also shown.}
\end{center}
\end{figure}

@@

\begin{Schunk}
\begin{Sinput}
R> javelin_table
\end{Sinput}
\begin{Soutput}
          throw1 throw2 throw3 throw4 throw5 throw6
Chopra     87.03  87.58  76.79      X      X  84.24
Vadlejch   83.98      X      X  82.86  86.67      X
Vesely     79.73  80.30  85.44      X  84.98      X
Weber      85.30  77.90  78.00  83.10  85.15  75.72
Nadeem     82.40      X  84.62  82.91  81.98      X
Katkavets  82.49  81.03  83.71  79.24      X      X
Mardare    81.16  81.73  82.84  81.90  83.30  81.09
Etelatalo  78.43  76.59  83.28  79.20  79.99  83.05
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> javelin_vector <- attemptstable2supp3(javelin_table,
+        decreasing = TRUE, give.supp = FALSE)
R> options(width = 60)
R> javelin_vector
\end{Sinput}
\begin{Soutput}
   Chopra    Chopra  Vadlejch    Vesely     Weber     Weber 
    87.58     87.03     86.67     85.44     85.30     85.15 
   Vesely    Nadeem    Chopra  Vadlejch Katkavets   Mardare 
    84.98     84.62     84.24     83.98     83.71     83.30 
Etelatalo     Weber Etelatalo    Nadeem  Vadlejch   Mardare 
    83.28     83.10     83.05     82.91     82.86     82.84 
Katkavets    Nadeem    Nadeem   Mardare   Mardare   Mardare 
    82.49     82.40     81.98     81.90     81.73     81.16 
  Mardare Katkavets    Vesely Etelatalo    Vesely Katkavets 
    81.09     81.03     80.30     79.99     79.73     79.24 
Etelatalo Etelatalo     Weber     Weber    Chopra Etelatalo 
    79.20     78.43     78.00     77.90     76.79     76.59 
    Weber  Vadlejch    Nadeem  Vadlejch    Chopra    Vesely 
    75.72        NA        NA        NA        NA        NA 
   Chopra Katkavets  Vadlejch    Vesely    Nadeem Katkavets 
       NA        NA        NA        NA        NA        NA 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> javelin <- ordervec2supp3(v = names(javelin_vector)[!is.na(javelin_vector)])
\end{Sinput}
\end{Schunk}

@@

\begin{figure}[htbp]
  \begin{center}
\begin{Schunk}
\begin{Sinput}
R> (mj <- maxp(javelin))
\end{Sinput}
\begin{Soutput}
   Chopra Etelatalo Katkavets   Mardare    Nadeem  Vadlejch 
   0.0930    0.0482    0.0929    0.1173    0.1730    0.3206 
   Vesely     Weber 
   0.1140    0.0409 
\end{Soutput}
\begin{Sinput}
R> dotchart(mj, pch = 16,xlab="Estimated Bradley-Terry strength")
\end{Sinput}
\end{Schunk}
\includegraphics{jss_hyper3-testthejav}
\caption{Maximum \label{javelinbradleyterry} likelihood estimate for javelin throwers' Bradley-Terry strengths.}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics{jss_hyper3-plottheloglikcont}
\caption{Profile \label{profliklogcont} likelihood for log-contrast
$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$.  Null
of $p_{\mbox{\tiny Vad}}=p_{\mbox{\tiny Ves}}$ indicated with a dotted
line, and two-units-of-support limit indicated with horizontal lines;
thus the null is not rejected.}
\end{center}
\end{figure}

@@

\begin{Schunk}
\begin{Sinput}
R> constructor_2021_table[, 1:9]
\end{Sinput}
\begin{Soutput}
   Constructor BHR EMI POR ESP  MON AZE FRA STY
1         Merc   1   2   1   1    7  12   2   2
2         Merc   3 Ret   3   3  Ret  15   4   3
3         RBRH   2   1   2   2    1   1   1   1
4         RBRH   5  11   4   5    4  18   3   4
5      Ferrari   6   4   6   4    2   4  11   6
6      Ferrari   8   5  11   7 DNSP   8  16   7
7           MM   4   3   5   6    3   5   5   5
8           MM   7   6   9   8   12   9   6  13
9           AR  13   9   7   9    9   6   8   9
10          AR Ret  10   8  17   13 Ret  14  14
11         ATH   9   7  10  10    6   3   7  10
12         ATH  17  12  15 Ret   16   7  13 Ret
13         AMM  10   8  13  11    5   2   9   8
14         AMM  15  15  14  13    8 Ret  10  12
15          WM  14 Ret  16  14   14  16  12  17
16          WM  18 Ret  18  16   15  17  18 Ret
17        ARRF  11  13  12  12   10  10  15  11
18        ARRF  12  14 Ret  15   11  11  17  15
19          HF  16  16  17  18   17  13  19  16
20          HF Ret  17  19  19   18  14  20  18
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> const2020 <- ordertable2supp3(constructor_2020_table)
R> const2021 <- ordertable2supp3(constructor_2021_table)
R> options(digits = 4)
R> maxp(const2020, n = 1)
\end{Sinput}
\begin{Soutput}
   ARRF     ATH Ferrari      HF    Merc      MR       R 
0.04530 0.06807 0.06063 0.02623 0.37783 0.10026 0.09767 
   RBRH  RPBWTM      WM 
0.12072 0.08055 0.02273 
\end{Soutput}
\begin{Sinput}
R> maxp(const2021, n = 1)
\end{Sinput}
\begin{Soutput}
    AMM      AR    ARRF     ATH Ferrari      HF    Merc 
0.05942 0.07543 0.06238 0.05611 0.16939 0.02023 0.19395 
     MM    RBRH      WM 
0.14126 0.18334 0.03848 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> H <- (
+       psubs(constructor_2020, "Merc", "Merc2020") +
+       psubs(constructor_2021, "Merc", "Merc2021")
+      )
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> options(digits = 4)
R> samep.test(H, c("Merc2020", "Merc2021"))
\end{Sinput}
\begin{Soutput}
	Constrained support maximization

data:  H
null hypothesis: Merc2020 = Merc2021
null estimate:
     AMM       AR     ARRF      ATH  Ferrari       HF 
 0.04239  0.05413  0.04677  0.04374  0.07568  0.02323 
Merc2020 Merc2021       MM       MR        R     RBRH 
 0.13903  0.13903  0.09016  0.07944  0.07475  0.10024 
  RPBWTM       WM 
 0.06235  0.02905 
(argmax, constrained optimization)
Support for null:  -1189 + K

alternative hypothesis:  sum p_i=1 
alternative estimate:
     AMM       AR     ARRF      ATH  Ferrari       HF 
 0.03766  0.04824  0.04333  0.04060  0.07036  0.02132 
Merc2020 Merc2021       MM       MR        R     RBRH 
 0.23135  0.09216  0.07893  0.07973  0.07455  0.09322 
  RPBWTM       WM 
 0.06177  0.02679 
(argmax, free optimization)
Support for alternative:  -1184 + K

degrees of freedom: 1
support difference = 4.722
p-value: 0.002119 
\end{Soutput}
\end{Schunk}

@@

\bibliography{jss4832}
\end{document}
